Literature Report on DP-means

### 2015-06-08 stat302/paper_report.tex
Started on paper report; finished overview

\documentclass[11pt]{article}
\usepackage[paper=letterpaper, margin=.5in]{geometry}
\pdfpagewidth 8.5in
\pdfpageheight 11in
\setlength\parindent{0in}

@@

\newcommand{\E}{\mathrm{E}}
\newcommand{\p}{\mathrm{P}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\ST}{mbox{ s.t. }}

@@

\newcommand{\st}{ \; \big | \:}
\newcommand{\car}[1]{\left\vert #1 \right\vert}

@@

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hwhead{STAT 347 Homework 4}
\begin{itemize}
    \item[1.]
        \begin{itemize}
            \item[(i)]
                After calculating the mean vector and the two covariance matricies, I found that
                \[ 
                    \hat\mu = \left[\begin{array}{c} 5.843 \\ 3.057 \\ 3.758 \\ 1.199 \\ \end{array}\right], \quad
                    S_b = \left[\begin{array}{cccc}  
                              31.606 & -9.976 & 82.624 & 35.640 \\ 
                              -9.976 & 5.672 & -28.620 & -11.466 \\ 
                              82.624 & -28.620 & 218.551 & 93.387 \\ 
                              35.640 & -11.466 & 93.387 & 40.207 \\ 
                        \end{array}\right]
                    \]
                and
                \[
                    S_w = \left[\begin{array}{cccc}
                            0.265 & 0.093 & 0.168 & 0.038 \\ 
                            0.093 & 0.115 & 0.055 & 0.033 \\ 
                            0.168 & 0.055 & 0.185 & 0.043 \\ 
                            0.038 & 0.033 & 0.043 & 0.042 \\
                        \end{array}\right]
                \]
            \item[(ii)]
                The eigenvalues of $S_w^{-1}S_b$ are 
                \[\lambda = 2366.107,\; 20.976, \; 0.000, \; 0.000\]
                This tells us that variation among blocks largely lies on a 1 or 2 dimmensional manifold. In particular, it can overwheminly be captured as occuring on a line. That $S_w^{-1}S_b$ follows from $S_w$ being positive definite. Due to this, we can say $S_w=BB$ For some positive definite matrix $B$. Then,
                \begin{align*}
                    S_w^{-1}S_b     &= BBS_b \\
                    B^{-1}S_w^{-1}S_bB &= BS_bB
                \end{align*}
                $BS_bB$ must be symmetric and thus have real eigenvalues. Since I have established $S_w^{-1}S_b$ is similar to a $BS_bB$, it too must have real eigenvalues. Since the eigenvalues are real, so are the eigenvectors, H. However, the $S_w^{-1}S_b$ need not be symmetric, so nothing is forcing $H$ to be orthogonal. Let $h_1$ be the first eigenvector. The portion of the variance due to the clustering in $Yh_1$ is smaller than for any other linear combination.
            \item[(iii)]
                To show that $\E(S_w)=\Sigma$, we observe
                \begin{align*}
                    \Sigma &= \E\left[\left(Y_i-\E[Y_i\st b_i]\right)'\left(Y_i-\E[Y_i\st b_i]\right)\right] \\
                           &= \E\left[\frac{1}{n-\car{B}}\sum_{i=1}^n\left(Y_i-\frac{1}{n_b}\sum_{j\in b} Y_j\right)'\left(Y_i-\frac{1}{n_b}\sum_{j\in b} Y_j\right)     \right] \\
                           &= \E\left[Y'Y + Y'\bar B Y\right] \\
                           &= \E\left[Y'(I+\bar B)Y\right] \\
                           &= \E\left[S_w\right] 
                \end{align*}
                Now, to show the result for $\E(S_b)$
                \begin{align*}
                    (\car{B}-1)S_b &= \sum(Y_iY_i') - n\bar Y\bar Y' - Y'(I-J)Y  \\
                    (\car{B}-1)S_b &= \sum(Y_iY_i'-\bar Y\bar Y') - Y'(I-J)Y  \\
                    (\car{B}-1)S_b &= \sum(Y_iY_i'-\frac{1}{n}\sum_b n_b\bar Y_b\bar Y_b') - Y'(I-J)Y  \\
                    (\car{B}-1)S_b &= \sum_b\left(\frac{1}{n}\sum_{i\in b}Y_iY_i'- \bar Y_b\bar Y_b'\right) - Y'(I-J)Y  \\
                    \E(S_b) &= \Sigma +\theta\Sigma\frac{n^2-\sum n^2_b}{n(\car{B}-1)}
                \end{align*}
                Using the result above, we can estimate the variance ratio based on 
                \[ \E\left[S_b\right] = \E\left[S_w\right]\left(1+\theta\frac{n^2-\sum n_b^2}{n(\car{B}-1}\right) \]
                Allowing us to derive an estimator
                \begin{align*}
                    tr(S_b) &= tr(S_w)\left(1+\hat\theta\frac{n^2-\sum n_b^2}{n(\car{B}-1)}\right) \\
                    \frac{tr(S_b)}{tr(S_w)} &=1+\hat\theta\frac{n^2-\sum n_b^2}{n(\car{B}-1)} \\
                    \frac{tr(S_b)}{tr(S_w)}-1 &=\hat\theta\frac{n^2-\sum n_b^2}{n(\car{B}-1)} \\
                    \hat\theta &= \frac{n(\car{B}-1)}{n^2-\sum n_b^2}\left(\frac{tr(S_b)}{tr(S_w)}-1\right) 
                \end{align*}
                Using this estimator, I get an estimate of $\hat\theta = 9.727$ on the original scale, and $\hat\theta = 18.210$ when the log is taken of the data.
            \item[(iv)]
                Calculating the predictive probabilities with $\lambda=1$ and $\theta=10$, I got this result:
                \FloatBarrier
                \input{hw4/A1_iv}
                \FloatBarrier
            \item[(v)]
                Now, with $\lambda=1$ and $\theta=5$, I got
                \FloatBarrier
                \input{hw4/A1_v_5}
                \FloatBarrier
                And with  $\lambda=1$ and $\theta=50$, I got
                \FloatBarrier
                \input{hw4/A1_v_50}
                \FloatBarrier
            The predictions do seem to be somewhat sensitive to $\theta$. In general, $\theta$ effects the relative probabilities of the existing clusters only slightly, but has a large effect on the probability of a new cluster.
            \item[(vi)]
                As $\theta\to0$
                \[ n_b\, \phi_4\!\left(y(u') - \frac{\mu+n_b\theta \bar y_b}{1+n_b\theta};\Sigma\left(1+\frac{\theta}{1+n_b\theta}\right)\right) \to n_b\, \phi_4\!\left(y(u') - \mu;\Sigma\right) \]
                and 
                \[ \lambda\, \phi_4\!\left(y(u') - \mu;\Sigma\left(1+\theta\right)\right) \to \lambda\, \phi_4\!\left(y(u') - \mu;\Sigma\right) \]
                This means that, since $\phi_4\!\left(y(u') - \mu;\Sigma\right)$ is a common factor,
                \[ \p(u' \mapsto b\st ...) \to \begin{cases} \frac{n_b}{\lambda+n} & b\in B \\
                                                         \frac{\lambda}{\lambda+n} & b = \emptyset 
                                                 \end{cases} \]
                Which is the Ewens process and independent of the feature vector. On the other hand, if $\theta\to\infty$, we see that
                \[ n_b\, \phi_4\!\left(y(u') - \frac{\mu+n_b\theta \bar y_b}{1+n_b\theta};\Sigma\left(1+\frac{\theta}{1+n_b\theta}\right)\right) \to n_b\, \phi_4\!\left(y(u') - \bar y_b;\Sigma\left(1+\frac{1}{n_b}\right)\right) \]
                and 
                \[ \lambda\, \phi_4\!\left(y(u') - \mu;\Sigma\left(1+\theta\right)\right) \to 0 \]
                so the predictive probability for the new class tends to zero and 
                \[ \p(u' \mapsto b \st b\in B ...) \propto n_b\phi_4\!\left(y(u') - \bar y_b;\Sigma\right) \]
                which is the Bayes optimal solution based on the Fisher discriminant and class frequencies based on sample frequencies. 
            \item[(vii)]
                Making the scatter plot of the projection:
                \begin{center}
                    \includegraphics[width=15cm]{hw4/A1_vii} 
                \end{center}
                Looking at this plot, its unsurprising that none of the values we predicted had high probabilities for both Setosa and another existing cluster; it is far more distinct from the other two than they are from each other. If its probability is of similar magnitude to one of the other clusters, then that means that the point is in the big gap in the middle, and the most likely prediction is probably a new cluster. Now, predicting $\bar y$, I got 
                \FloatBarrier
                \input{hw4/A1_vii}
                \FloatBarrier
                The overall mean is much closer to the Versicolor mean than the other means, so these probabilities are very skewed towards Versicolor.
            \item[(viii)]
                Here is the alternate plot of the data
                \begin{center}
                    \includegraphics[width=15cm]{hw4/A1_viii} 
                \end{center}
                This is a plot of the data's projection into the first two principal components of the data matrix (though without subtracting out the mean or normalizing). We are still capturing the majority of the variation in the data; we're just using a different basis.
            \item[(ix)]
                One might want to use a log scale since all the measurements by their nature must be positive, which defies the normality assumption. When one makes predictions based on the log scale, these are the new probabilities.
                \FloatBarrier
                \input{hw4/A1_ix}
                \FloatBarrier
                Besides the second point, the predictions are quite similar as before.
        \end{itemize}
    \item[5.]
        \begin{itemize}
            \item[(i)] {\bf Show this is a probability distribution:}
                This follows from this manipulation
                \begin{align*}
                    \p(T=t;m,n) &= \frac{\prod_i n_i! \prod_j m_j!}{n_*! \prod_{ij} t_{ij}!} \\
                                &= \frac{n_*!\prod_i n_i! \prod_j m_j!}{n_*!n_*! \prod_{ij} t_{ij}!} \\
                                &= \frac{\prod_i n_i!}{n_*!}\frac{\prod_j m_j}{n_*!}\frac{n_*!}{ \prod_{ij} t_{ij}!} \\
                                &= \frac{{{n_*}\choose {t_{11},...,t_{cr}}}}{{{n_*}\choose {n_1,...,n_r}}{{n_*}\choose {m_1,...,m_c}}}
                \end{align*}
                The numerator is the number of ways to fill fill each cell with the desired $t_{ij}$ from $n_*$ observations, and the denominator is the number of all possible tables which satisfy the marginal distributions. Thus, it is clear that summing the probability over all possible $T$ which satisfy the marginals will yield 1, making this a proper probability distribution.
            \item[(ii)] {\bf Show $T=\tilde A'P\tilde B$ has a hypergeometric distribution:} Since $PB$ is just a permutation of the rows of $B$, the column sums will remain the same. This means that the marginals for the new factor mapping to $PB$ is the same as for the factor for $B$. In turn, the resulting table $A'PB$ will have the same marginals in both the rows and columns, so clearly there is the proper conditioning. Further, each permutation is chosen with probability $\frac{1}{n_*!}$. However, most of these permutations result in the same table. In particular, if one permutes the values of $B$ matched with each particular value of $A$, then the table is the same. There are $\prod_i n_i!$ such permutations. Similarly, permutations of a particular value of $B$, of which there are $\prod_j m_j$ will also result in the same table. Then, there are $\prod_{ij}t_{ij}!$ permutations that do both of these things at once. It follows that there are 
                \[ \frac{\prod_i n_i! \prod_j m_j!}{\prod_{ij} t_{ij}!} \]
                permutations total that result in the same table. Thus, we conclude that the probability under this model of drawing a particular table is
                \[\frac{\prod_i n_i! \prod_j m_j!}{n_*! \prod_{ij} t_{ij}!}\]
                Making it a hypergeometric distribution.
            \item[(iii)] {\bf Compute Pearson's chi-squared test and deviance statistic:} 
                I found that the nominal Pearson Chi-square statistic is 109.883, and the deviance statistic is
                \[ 2\sum_{i,j}t_{ij}\left(\log(t_{ij})-\log\left(n_*\frac{m_j}{n_*}\frac{n_i}{n_*}\right)\right) = 130.599 \]
            \item[(iv)] {\bf Estimate the p-value by simulation:} 
                Over $10000$ simulations, I found that the Perason's chi-squared test statistic was exceeded 7723 times, and the deviance statistic was exceeded 6689 times. These correspond to p-values of .772 and .669 respectively. Below I've plotted both sets of bootstraps versus the density of the expected chi-square with 121 degrees of freedom: \\
                \includegraphics[width=9cm]{hw4/A5_chi} 
                \includegraphics[width=9cm]{hw4/A5_dev} 
                The Pearson statistic is extremely close the Chi-sq, as one would hope for. However, the deviance statistic is noticeably shifted to the right. This 
            \item[(v)] {\bf Aggregate the data by birth death month difference:} 
                After aggregating the expected and actual birth and death combinations by the difference in month, this is the resulting table:
                \FloatBarrier
                \input{hw4/A5_sum_diff}
                \FloatBarrier
                We seem to see fewer than expected deaths in the months leading up to someone's birthday, and more deaths than expected afterwards. Using the Pearson Chi-squared test to evaluate this, I get a p-value of $.027$, which is evidence, though not overwhelming, supporting that this is statistically significant.

            \item[(vi)] {\bf Explain how to extend the hypergeometric simulation to three dimensions where each one dimensional table fixed:} 
                For each cell, the expected count in each cell under the null should still be the product of the marginal probabilities times the total $n$. Thus, for for the indicator matrices $\tilde A$, $\tilde B$, $\tilde C$, we can simulate a random null matrix fitting the marginals using random permutation matrices $P_1$ and $P_2$. A cell of the table is 
                \[ T_{i,j,k} = \sum_{m=1}^n \tilde A_{mi} (P_1\tilde B)_{mj} (P_2\tilde C)_{mk} \]
                with expectation, under the null, of
                \[ \E[T_{i,j,k}\st H_o] = n\frac{\|\tilde A_i\|_1}{n}\frac{\|\tilde B_j\|_1}{n}\frac{\|\tilde C_k\|_1}{n}\]
                From here, the simulation is the same, bootstrapping a test statistic over many simulated tables and comparing it to the observed value.
            \item[(vii)] {\bf Explain how to extend the hypergeometric simulation to three dimensions with two-dimension tables fixed:} 
                Here, the null is that the cells are conditionally independent. In other words,
                \begin{align*}
                    \p(A_m=i,B_m=j,C_m=k) &= \p(A_m=i,B_m=j\st C_m=k)\p( C_m=k) \\
                                          &= \p(A_m=i\st C_m=k)\p(B_m=j\st C_m=k)\p( C_m=k) \\
                \end{align*}
                This can be tested by the same method as with two way tables, just induced on all the subtables $T_{*,*,k}$ at the same time. For each subset of $A_k$ and $B_k$, which are the values of $A$ and $B$ for which $C=k$, draw a random permutation $P$, and then $\tilde A_k' P \tilde B_k$ should be hypergeometric. A test statistic is then generated on all the two way tables for all $k$ by comparing the seen cells $T_{i,j,k}$ to
                \[ \E(T_{i,j,k}) = \frac{1}{u_k}\sum_{i'} T_{i',j,k}\sum_{j'}T_{i,j',k} \]
                Where $u_k$ is the number of times level $k$ of $C$ occurs.
        \end{itemize}
\end{itemize}

>>>>

\documentclass[11pt]{article}
\usepackage[paper=letterpaper, margin=.75in]{geometry}
\linespread{1.5}
\pdfpagewidth 8.5in
\pdfpageheight 11in

@@

\newcommand{\E}{\mathrm{E}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\ST}{mbox{ s.t. }}

@@

\newcommand{\st}{ \; \big | \:}

@@

\begin{document}
\title{Literature Report on DP-means}
\author{Aaron Maurer}
\date{STAT 302, Spring Quarter 2015}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
For this report, I chose to read "Revisiting k-means: New Algorithms via Bayesian Nonparametrics"\footnote{Brian Kulis and Michael I. Jordan, ``Revisiting k-means: New Algorithms via Bayesian Nonparametrics'', \textit{CoRR} (2011): http://arxiv.org/abs/1111.0352}. This article introduces a alternative to the k-means clustering algorithm called DP-means, where instead of specifying a number of clusters at the start, then working to optimize their fit, new clusters are introduced when a point in the data set is too far from one of the existing clusters. This in itself is a `hard' clustering algorithm, which outputs an assignment to clusters, rather than a posterior distribution one might normally expect from a Bayesian method. However, this paper demonstrates that this algorithm is the limit of a Bayesian model where the total distribution is a mixture of Gaussian distributions, with the number of components coming from a Dirichlet process. This is the source of the name of the algorithm: Dirichlet process means or DP-means. \par
In addition to a description and derivation of this algorithm, the paper offers a few extensions and simulations. The most interesting of theses is a hierarchical clustering method where, over multiple data sets of the same variables, local clusters are simultaneously fit on each data set so as to match a set of global clusters across all of the data sets. This is again a hard clustering algorithm, but it can be derived as well in a Bayesian fashion by taking the limit when the parameters of the Dirichlet process which generates the components in each data set arises from a prior global Dirichlet process. The next extension shows how DP-means extends to spectral clustering; where one would spectrally cluster using k-means by performing k-means on the first k eigenvectors of a similarity matrix for the data, instead with DP-means one takes all eigenvectors where the eigenvalues are above a given threshold, then cluster via DP-means using that same threshold to determine when to add additional components. Similarly, the author also shows that DP-means can be extended to graph cut problems. The paper concludes with a few simulations demonstrating the effectiveness of DP-means and its multiple data set extension. \par
This article offers obvious extensions of how the classical Bayesian statistics we learned can be extended to a machine learning algorithm. The prior Dirichlet process is used to guide the posterior number of components. The method for fitting the model is the limit of a Gibbs sampling algorithm. Finally, the multiple data set version of DP-means is a hierarchical model, based on the exchangeability of the data sets. All together, this forms an interesting competitor to k-means, being similarly easy to compute, but built on classic Bayesian principles.

\section{Paper Contents}
\subsection{DP-Means Algorithm}
